Epistemic Semantics in Guarded String Models

### 2020-09-25 tex/sdraft.tex
tighten section 4

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics \citep{kaplan1994regular,beesley2003finite}.  Current
implementations allow for the definition of functions that have the status of defined
operations on regular sets and relations \citep{hulden2009foma,karttunen2010update}.
Such definitions are used here to construct of a model for epistemic KAT inside the
finite state calculus.  The space of worlds is a set of ordinary strings.  Bit
sequences (sequences of 0's and 1's) are used for atoms, and these alternate with
event symbols in the encoding of a world. \ex{gs1fst} gives the encoding of worlds $w_2$ and $w_3$
from the example.  A string is a finite sequence of symbols, and {\tt 0},
{\tt 1}, {\tt u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
multi-character symbols that are found in implementations of the finite state
calculus. The multi-character symbol {\tt a0} is an element of the alphabet that has
no connection with the element {\tt a}.

@@

Terms in the finite state calculus are interpreted as sets of strings,
or for relational terms, as relations between strings.
Computationally, the sets and relations are represented by finite
state acceptors.  As used here, a program in the Fst language of the
finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and
functions (defined as macros) mapping one or more regular sets or
relations to a regular set or relation.  The finite state calculus has
a product operation of string concatenation raised to sets.
Concatenation of strings with atoms (Boolean vectors) at both ends has
the effect of doubling atoms at the juncture, and does not enforce
matching of atoms at the juncture.  Therefore KAT product can not be
identified with product in the finite state calculus. Instead, KAT
product and KAT Kleene star are defined operations, see Figure
\ref{fstfigure1}.  The binary product operation {\tt Cn} and the unary
Kleene star operation {\tt Kst} combine strings in the string algebra,
remove strings with non-matching atoms, and then delete the second of
two tests to create well-formed guarded strings.  Matching of atoms is
enforced with the set {\tt Wf0}, which is the set of strings that do
not contain non-matching Boolean vectors.  The containment operator
(expressed by the dollar sign) and complement (expressed by tilde) are
operators of the finite state calculus.  The set of non-matching
sequences of atoms {\tt UnequalStPair} is defined by a finite
disjunction.  Deletion is accomplished by a re-write rule in the
finite state calculus, which is a notation for defining regular
relations by contextually constrained substitutions.  In this case, is
{\tt Squash} is a regular relation that deletes an atom (a sequence of
0's and 1's of a certain length) when it follows an atom.\footnote{
  This is a non-equal length regular relation. The finite state
  calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted
  in that the complement and set difference for non-equal length
  relations is not defined.  The epistemic alterative relations that
  are defined in Figure -- are equal-length relations.  }

@@

An event symbol such as $a_1$ (Amy looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.

@@

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).  In
the construction in Fst, the definition of world alternatives takes a
simple form. Where $R_a$ is the relation on decorated events for agent
$a$, the the corresponding relation on worlds in is the Kleene closure
of $R_a$.  Where $R$ and $S$ are relations, the concatenation product
of $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$,
where $\pair{x_1}{y_1}$ is in relation $R$, and $\pair{x_2}{y_2}$ is
in relation $S$.  The Kleene closure of relation $R$ is \(\cup_{n \geq
  0}R^n\), where $R^n$ is the $n$-times concatenation product of $R$
with itself (the $0$-times concatenation product is $\sem{ 1
}^{\varphi,\mathcal E}$). This is an operation in the finite
state calculus.  Figure \ref{relclosure} defines the corresponding
operation in the guarded string algebra.  The epistemic alternative
relation on worlds for an agent is then defined as the concatenation
closure of the event alternative relation for the agent.

>>>>

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic KAT inside the finite state
calculus. We describe our translation from Epik terms to Fst programs
here.

First, the space of worlds is a set of ordinary (as opposed to
guarded) strings.  Bit sequences (sequences of $0$'s and $1$'s) encode
atoms, as before, these alternate with event symbols to encode a
world. \ex{gs1fst} gives the encoding of worlds $w_2$ and $w_3$ from
the example. The valuation of $h$ preceeds the valuation of $t$, so
$1~0$ indicates the valuation $h\bar{t}$.

%% unnecessary detail
%% A string is a finite sequence of symbols, and {\tt 0}, {\tt 1}, {\tt
%%   u}, and {\tt d}, are symbols.  {\tt a0} and {\tt b0} are
%% multi-character symbols that are found in implementations of the
%% finite state calculus. The multi-character symbol {\tt a0} is an
%% element of the alphabet that has no connection with the element {\tt
%%   a}.

@@

Terms in the finite state calculus are interpreted as sets of strings,
or for relational terms, as relations between strings.
Computationally, the sets and relations are represented by finite
state acceptors.  As used here, a program in the Fst language of the
finite state calculus is a straight-line program that defines a
sequence of constants naming sets, constants naming relations, and
functions (defined as macros) mapping one or more regular sets or
relations to a regular set or relation.

Now, translating the Epik terms $0$, $1$, $b$, and $e$ are
straightforward: we simply convert the atoms as previously described,
ensuring to decorate the events $e$ with their compatible atoms
first. For example $a_1$ becomes $\{01a_101\}$ Fst also has built-in notions of union (\texttt{|}),
intersection (\texttt{\&}), and complement (\texttt{\~}). Translating
the the remaining operators requires more work.

Implementing fusion product using Fst's standard string concatenation
(denoted by juxatposition $\texttt{X} \texttt{Y}$) requires more
care. Naively concatenating strings with atoms (Boolean vectors) at
both ends doubles the atoms at the juncture, and does not enforce the
requisite atom equality.

To implement the fusion product, we define the binary product
operation \texttt{Cn}, which combines strings in the underlying string
algebra, remove strings with non-matching atoms, and then delete the
second of two tests to create well-formed guarded strings (see Figure
\ref{fstfigure1}). The set {\tt Wf0} describes pairs of equal atoms as
the complement of set of unequal atom pairs, implemented using Fst's
containment operator \texttt{\$}. Then. to delete the second of the
two (matching) pairs, the \texttt{Squash} operator uses Fst's rewrite
notation to delete atoms (\texttt{St}) that are preceded by another
atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used
  with relation composition and relation domain and co-domain.  They
  are restricted in that the complement and set difference for
  non-equal length relations is not defined.  The epistemic alterative
  relations that are defined in Figure -- are equal-length relations.
} The squash operator is applied via the relational composition
\texttt{.o.} and codomain \texttt{.l} operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

Kleene star is defined in a similar way (see Figure \ref{fstfigure1}),
unioning the set of atoms \texttt{St} with \texttt{Kpl} which uses
Fst's non-empty iteration operator \texttt{+} and normalizes the
result in the same way as \texttt{Cn}.

@@

%% An event symbol such as $a_1$ (Amy looks at heads) is in the KAT algebra a set of
%% bare events decorated with compatible tests on each side, $\{ 10 a_1 10
%% \}$ in this case.  This is a unit set rather than a guarded string, because elements
%% of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
%% events using {\em Kst}. The operation enforces compatibility of states, so that
%% $(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
%% Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
%% KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$.

@@

The relation on decorated events needs to be generalized to a relation
of worlds.  The principle for this is that an epistemic alternative to
a world of the form $we$ is a world of the form $vd$, where $v$ is a
world-alternative to $w$, $d$ is an event-alternative to $e$, and $vd$
is defined (i.e.  the world alternative $v$ satisfies the
pre-conditions of the event alternative $d$).  This principle is found
in earlier literature (Moore 198x, Baltag, Moss and Solecki 20xx).

In the construction in Fst, the definition of world alternatives takes
a simple form. Where $R_a$ is the relation on decorated events for
agent $a$, the corresponding relation on worlds in is the Kleene
closure of $R_a$.  Where $R$ and $S$ are relations, the concatenation
product of $R$ and $S$ is the set of pairs of the form
$\pair{x_1x_2}{y_1y_2}$, where $\pair{x_1}{y_1}$ is in relation $R$,
and $\pair{x_2}{y_2}$ is in relation $S$.  The Kleene closure of
relation $R$ is \(\cup_{n \geq 0}R^n\), where $R^n$ is the $n$-times
concatenation product of $R$ with itself (the $0$-times concatenation
product is $\sem{ 1 }^{\varphi,\mathcal E}$). This is an operation in
the finite state calculus.  Figure \ref{relclosure} defines the
corresponding operation in the guarded string algebra.  The epistemic
alternative relation on worlds for an agent is then defined as the
concatenation closure of the event alternative relation for the agent.
